Automating Free Logic in HOL, with an Experimental Application in Category Theory

### 2018-10-12 FreeLogic/2018-JAR/article.tex
modfied

@@

\section{Preliminaries}
\label{sec:preliminaries}
\subsection{Free Logic}
Free logic (respectively
inclusive logic) 
\cite{Lambert60,Scott67,lambert02:_free_logic,sep-logic-free} refers to a class of logic
formalisms that are free of basic existence assumptions regarding the
denotation of terms. Remember that terms in e.g. traditional classical
and intuitionistic predicate logics always denote an (existing) object
in a given (non-empty) domain $\D$, and that $\D$ is also exactly the set the
quantifiers range over. In free logic these
basic assumptions are abolished. Terms do still denote objects in a
(non-empty) domain $\D$, but a (possibly empty) set $\ex\subset \D$ is
chosen to characterize the subdomain of ``\emph{existing}'' resp. ``\emph{defined}''
objects in $\D$. Quantification is
now restricted to set $\ex$ of existing/defined objects only.

@@

The sources of our experiments, as discussed in
\cite{ArXiv}, are available at
\url{http://christoph-benzmueller.de/papers/2017-arXiv.zip}. Note that
in these sources we added neither the definite description operator
nor the designated undefined object $\hol{\star}$, since both were not
required in this particular application. A related Isabelle/HOL
encoding of the above theory with description has been presented and
discussed in \cite{ICMS}.

@@

\begin{figure}[htp]
 \includegraphics[width=\textwidth]{FFOL-Pic1}
 \caption{Isabelle/HOL encoding of $\FFOL$ (without description)\label{pic1}}
\end{figure}

\begin{figure}[htp]
 \includegraphics[width=\textwidth]{FFOL-Pic2}
 \caption{Encoding of Axioms Set V in Isabelle/HOL utilizing the
   embedded logic $\FFOL$ \label{pic2}}
\end{figure}

@@

\subsection{Remark on the Experiments}  \label{sec-remarks-experiments}
\marginpar{Include remark on SMT calls and the SMT tactic (see Reviewer 1)}
\marginpar{Include remark on Sledgehammer and how it works (see Reviewer 1)}

All proofs above and all proofs in the rest of this paper (unless
stated otherwise) have been obtained fully automatically in very
reasonable time (typically just a few seconds) with the
Sledgehammer tool in Isabelle/HOL. This tool interfaces to prominent
first-order automated theorem provers such as CVC4 \cite{CVC4}, Z3
\cite{Z3}, E \cite{E} and Spass \cite{Spass}.  Remotely, also
  provers such as Vampire \cite{Vampire}, or the higher-order provers
  Satallax \cite{Satallax} and LEO-II \cite{LEO} can be reached. For
  example, to prove axiom $E_{iii}$ from Axioms Set II, we have called
  Sledgehammer on all axioms of Axioms Set II. The provers then, via
  Sledgehammer, suggested to call trusted/verified tools in
  Isabelle/HOL with the exactly required dependencies they
  detected, in this case $C_{ii}$, $D_{ii}$, $E_{ii}$ and $S_{ii}$. With the provided dependency information the trusted tools
  in Isabelle/HOL were then able to reconstruct the external proofs on
  their own.  This way we obtain a verification of our claims in
  Isabelle/HOL, in which all the proofs have nevertheless been
  contributed by automated theorem provers.

\marginpar{Remove this part? SMT is now called as an oracle.}
  However, there are still
  a few calls of the \textsc{smt}-tactic contained in our document, 
   which we found are not entirely trivial to replace. In this sense, our
   Isabelle/HOL verification is modulo the correctness of the \textsc{smt} solvers
   CVC4 \cite{CVC4} and Z3 \cite{Z3}.

@@

\begin{table} \centering \normalsize
\begin{tabular}{ll}
\hline
\\
\multicolumn{2}{c}{Axioms Set FS-I:  Freyd and Scedrov in original notation (with issues)} \\
%\hline
\\
  $A1$ &  $\ex(x \circ y) \longleftarrow (x\Box \cong \Box y)$ \\ 
  $A2a$ & $((\Box x)\Box) \cong \Box x$ \\ 
  $A2b$ & $\Box (x\Box) \cong \Box x$ \\ 
  $A3a$ & $(\Box x) \circ x \cong x$ \\ 
  $A3b$ & $x \circ (x\Box) \cong x$ \\ 
  $A4a$ & $\Box(x \circ y) \cong \Box(x \circ (\Box y))$ \\ 
  $A4b$ & $(x \circ y)\Box \cong ((x\Box) \circ y)\Box$ \\ 
  $A5$   & $x \circ (y \circ z) \cong  (x \circ y) \circ z$ \\
\\
\hline
\\
\multicolumn{2}{c}{Axioms Set FS-II: Freyd and Scedrov in our notation (with issues)} \\
%\hline
\\
  $A1$  & $\ex (x \comp y) \longleftrightarrow \dom x \cong \cod y$ \\
  $A2a$ & $\cod(\dom x) \cong \dom x$ \\  
  $A2b$ & $\dom(\cod y) \cong \cod y$ \\  
  $A3a$  & $x \comp (\dom x) \cong x$ \\ 
  $A3b$ & $(\cod y) \comp y \cong y$ \\
  $A4a$ & $\dom(x \comp y) \cong \dom((\dom x) \comp y)$ \\ 
  $A4b$ & $\cod(x \comp y) \cong \cod(x \comp (\cod y))$ \\ 
  $A5$   & $x \comp (y \comp z) \cong  (x \comp y) \comp z$   \\
\\
\hline
\\
\multicolumn{2}{c}{Axioms Set VI: Freyd and Scedrov in our notation
  and corrected} \\
%\hline
\\
  $A1'$  & $\ex (x \comp y) \longleftrightarrow \dom x \simeq\cod y$ \\
  $A2a$ & $\cod(\dom x) \cong \dom x$ \\  
  $A2b$ & $\dom(\cod y) \cong \cod y$ \\  
  $A3a$  & $x \comp (\dom x) \cong x$ \\ 
  $A3b$ & $(\cod y) \comp y \cong y$ \\
  $A4a$ & $\dom(x \comp y) \cong \dom((\dom x) \comp y)$ \\ 
  $A4b$ & $\cod(x \comp y) \cong \cod(x \comp (\cod y))$ \\ 
  $A5$   & $x \comp (y \comp z) \cong  (x \comp y) \comp z$   \\
\\
\hline
\end{tabular}
\caption{The axioms set of Freyd and Scedrov in their and our
  notation, together with a proposed correction.\label{axioms-sets-2}}
\end{table}

@@

  The (constricted) inconsistency of Axioms Set FS-I, respectively
  Axioms Set FS-II, from Table \ref{axioms-sets-2} has been 
  detected first by the model finder Nitpick. When we asked Nitpick to
  generate a model with at least one non-existing object, it claimed
  that there is no such model. However, a model can still be
  constructed if we do not make any assumptions about non-existing
  objects. In fact, the model presented by Nitpick for this case
  consists of a single, existing morphism.

@@

Further work includes the continuation of our formalization studies in
category theory and the application of free logic to various other
mathematical domains, including, for example,  projective geometry.

>>>>

\usepackage{cite}

@@

\section{Preliminaries}
\label{sec:preliminaries}
\subsection{Free Logic}
Free logic (respectively
inclusive logic) 
\cite{Lambert60,Scott67,lambert02:_free_logic,sep-logic-free} refers to a class of logic
formalisms that are free of basic existence assumptions regarding the
denotation of terms.\footnote{Calculi for free logic are presented in \cite{sep-logic-free}; see also the references
therein.} Remember that terms in e.g. traditional classical
and intuitionistic predicate logics always denote an (existing) object
in a given (non-empty) domain $\D$, and that $\D$ is also exactly the set the
quantifiers range over. In free logic these
basic assumptions are abolished. Terms do still denote objects in a
(non-empty) domain $\D$, but a (possibly empty) set $\ex\subset \D$ is
chosen to characterize the subdomain of ``\emph{existing}'' resp. ``\emph{defined}''
objects in $\D$. Quantification is
now restricted to set $\ex$ of existing/defined objects only.

@@

The sources of our experiments as conducted here are available at
\url{http://christoph-benzmueller.de/papers/2018-JAR-sources.zip}.
These sources contain an embedding of full free logic in Isabelle/HOL,
that is, with $\star$ and definite description
(cf.~Fig.\ref{pic1}). In \cite{ArXiv}, $\star$ and definite
description were still avoided; they are in fact not really 
relevant for the conducted experiments.
% Note that
% in these sources we added neither the definite description operator
% nor the designated undefined object $\hol{\star}$, since both were not
% required in this particular application. A related Isabelle/HOL
% encoding of the above theory with description has been presented and
% discussed in \cite{ICMS}.

@@

\begin{figure}[htp]
 \includegraphics[width=\textwidth]{FFOL-Pic1}
 \caption{Isabelle/HOL encoding of $\FFOL$ (with $\star$ and definite description). \label{pic1}}
\end{figure}

\begin{figure}[htp]
 \includegraphics[width=\textwidth]{FFOL-Pic2}
 \caption{Encoding of Axioms Set V in Isabelle/HOL utilizing the
   embedded logic $\FFOL$; Axioms Set V is proven equivalent to Axioms
   Set IV. \label{pic2}}
\end{figure}

@@

\subsection{Remark on the Experiments}  \label{sec-remarks-experiments}
% \marginpar{Include remark on SMT calls and the SMT tactic (see Reviewer 1)}
% \marginpar{Include remark on Sledgehammer and how it works (see
%   Reviewer 1)}
% \marginpar{Add footnote on some issues with Sledgehammer, Z3 and Spass}

All proofs above and all proofs in the rest of this paper (unless
stated otherwise) have been obtained fully automatically in very
reasonable time (typically just a few seconds) with the
Sledgehammer tool in Isabelle/HOL (version Isabelle2017). This tool interfaces to prominent
first-order automated theorem provers such as CVC4 \cite{CVC4}, Z3
\cite{Z3}, E \cite{E} and Spass \cite{Spass}.  Remotely, also
  provers such as Vampire \cite{Vampire}, or the higher-order provers
  Satallax \cite{Satallax} and LEO-II \cite{LEO} can be reached. For
  example, to prove axiom $E_{iii}$ from Axioms Set II, we have called
  Sledgehammer on all axioms of Axioms Set II. The provers then, via
  Sledgehammer, suggested to call trusted/verified tools in
  Isabelle/HOL with the exactly required dependencies they
  detected, in this case $C_{ii}$, $D_{ii}$, $E_{ii}$ and $S_{ii}$. With the provided dependency information the trusted tools
  in Isabelle/HOL were then able to reconstruct the external proofs on
  their own.  This way we obtain a verification of our claims in
  Isabelle/HOL, in which all the proofs have nevertheless been
  contributed by automated theorem provers. For further information on
  the use and functioning of Sledgehammer we refer to the 
  literature \cite{Sledgehammer,SledgehammerTutorial}.

  In our experiments we have also made use of the Isabelle/HOL's
  \textsc{smt} method, which ``translates the conjecture and any
  user-supplied facts to the SMT solvers’ many-sorted first-order
  logic, invokes a solver, and (depending on the solver) either trusts
  the result or attempts to reconstruct the proof in Isabelle.''
  (\cite[p.~5]{Sledgehammer}).\footnote{Technical remark: We have
    selected CVC4 in our experiments as the default SMT solver, since
    we did run into errors when working with Z3. These errors can
    easily be reconstructed in the provided source files when
    switching back to Z3 as default.}  For quite some time the use of the \textsc{smt}
  method has been controversially discussed in the Isabelle/HOL
  community. However, since about a year, the \textsc{smt} method is
  now even eligible in submissions to the Archive of Formal
  Proofs. For the work presented here this community internal discussion is 
  of little relevance, since we could replace all uses of the\textsc{smt} method
  by alternative proofs (which we avoid in order to keep
  our formalisation concise).

% \marginpar{Remove this part? SMT is now called as an oracle.}
%   However, there are still
%   a few calls of the \textsc{smt}-tactic contained in our document, 
%    which we found are not entirely trivial to replace. In this sense, our
%    Isabelle/HOL verification is modulo the correctness of the \textsc{smt} solvers
%    CVC4 \cite{CVC4} and Z3 \cite{Z3}.

@@

\begin{table} \centering \normalsize
\begin{tabular}{ll}
\hline
\\
\multicolumn{2}{c}{Axioms Set FS-I:  Freyd and Scedrov in original notation (with issues)} \\
%\hline
\\
  $A1$ &  $\ex(x \circ y) \longleftrightarrow (x\Box \cong \Box y)$ \\ 
  $A2a$ & $((\Box x)\Box) \cong \Box x$ \\ 
  $A2b$ & $\Box (x\Box) \cong \Box x$ \\ 
  $A3a$ & $(\Box x) \circ x \cong x$ \\ 
  $A3b$ & $x \circ (x\Box) \cong x$ \\ 
  $A4a$ & $\Box(x \circ y) \cong \Box(x \circ (\Box y))$ \\ 
  $A4b$ & $(x \circ y)\Box \cong ((x\Box) \circ y)\Box$ \\ 
  $A5$   & $x \circ (y \circ z) \cong  (x \circ y) \circ z$ \\
\\
\hline
\\
\multicolumn{2}{c}{Axioms Set FS-II: Freyd and Scedrov in our notation (with issues)} \\
%\hline
\\
  $A1$  & $\ex (x \comp y) \longleftrightarrow \dom x \cong \cod y$ \\
  $A2a$ & $\cod(\dom x) \cong \dom x$ \\  
  $A2b$ & $\dom(\cod y) \cong \cod y$ \\  
  $A3a$  & $x \comp (\dom x) \cong x$ \\ 
  $A3b$ & $(\cod y) \comp y \cong y$ \\
  $A4a$ & $\dom(x \comp y) \cong \dom((\dom x) \comp y)$ \\ 
  $A4b$ & $\cod(x \comp y) \cong \cod(x \comp (\cod y))$ \\ 
  $A5$   & $x \comp (y \comp z) \cong  (x \comp y) \comp z$   \\
\\
\hline
\\
\multicolumn{2}{c}{Axioms Set VI: Freyd and Scedrov in our notation
  and corrected} \\
%\hline
\\
  $A1'$  & $\ex (x \comp y) \longleftrightarrow \dom x \simeq\cod y$ \\
  $A2a$ & $\cod(\dom x) \cong \dom x$ \\  
  $A2b$ & $\dom(\cod y) \cong \cod y$ \\  
  $A3a$  & $x \comp (\dom x) \cong x$ \\ 
  $A3b$ & $(\cod y) \comp y \cong y$ \\
  $A4a$ & $\dom(x \comp y) \cong \dom((\dom x) \comp y)$ \\ 
  $A4b$ & $\cod(x \comp y) \cong \cod(x \comp (\cod y))$ \\ 
  $A5$   & $x \comp (y \comp z) \cong  (x \comp y) \comp z$   \\
\\
\hline
\end{tabular}
\caption{The axioms set of Freyd and Scedrov in their and our
  notation, together with a proposed correction.\label{axioms-sets-2}}
\end{table}

@@

  The (constricted) inconsistency of Axioms Set FS-I, respectively
  Axioms Set FS-II, from Table \ref{axioms-sets-2} has been 
  detected first by the model finder Nitpick. When we asked Nitpick to
  generate a model with at least one non-existing object, it claimed
  that there is no such model. However, a model can still be
  constructed if we do not make any assumptions about non-existing
  objects.\footnote{For this we have to inactivate the axiom
    that postulates that $\star$ is an undefined/non-existing object.} In fact, the model presented by Nitpick for this case
  consists of a single, existing morphism.

@@

Further work includes the extension of our work towards an embedding of
free higher-order logic, the continuation of our formalization studies in
category theory and the application of free logic to various other
mathematical domains, including, for example,  projective geometry.
Regarding extensions towards free  higher-order logic some first steps
have already been taken \cite{W56,makarenko16:_autom_logik_logik_stufe}.
